Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-25 doc/response.tex

\jd{I would skip the following text.}

The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial effects.
The Laplace tranforms are only more useful for the intrinsic generation-interval distributions under the homogeneous assumption. 
The forward generation-interval distribution in a spatially hetergeneous environment is analytically intractable and therefore, the Laplace transforms no longer hold.
Given that our goal is to infer the initial forward generation-interval distribution, rather than the intrinsic distribution, we feel that the Laplace transforms are not necessarily helpful for the discussion.
In particular, the reviewer asks us to prove special cases of the Euler-Lotka equation, which has been proven previosuly as suggested by the references, using Laplace transforms; 
we do not think this is relevant to the discussion of our paper.

\revtext{How to assess if the gamma distribution is a good choice as a
model? Run some kind of goodness-of-fit tests?}

These are hard issues, but beyond the scope of this paper. Using exponentials is routine in many applications. Using gammas (or Erlangs) generalizes this, and is also commonly done. 
The gamma distribution has been previously tested for many diseases. 
In our case, Figure 4 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as cuss potential caveats in assuming a shape of the generation-interval distribution in Discussion.

@@

\newtext{We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity;
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).}

\jd{I am not sure what the last part could mean, except that some biases don't turn out as bad as others -- they can't cancel exactly, so we shouldn't say unbiased estimates... Am I missing something?}

@@

\revtext{How does the likelihood
function look like? Since the "observed" generation-intervals (assuming
observable) do not arise from a designed random experiment, but observational data, I expect the likelihood function should reflect the observation
scheme (e.g. retrospective)? Please be explicit.
}

The population-level method does not take into account the observation scheme or the infection process. 
We simply fit the gamma distribution to all realized generation intervals (i.e., the aggregated distribution).
We have clarified the text:

\newtext{The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom):
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau).
\end{equation}
In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a by fitting a gamma distribution to the observed values by maximum likelihood.
The inferred initial forward generation-interval distribution also follows a gamma distribution with the same shape, but with mean $\alpha/(\alpha/\bar G - r)$.
We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).}

@@

\revtext{
The paper focuses on estimating the generating-interval distribution. There are
several generating-intervals and distributions mentioned in this paper, such as
$g(\tau)$; $b_t(\tau)$; $c_t(\tau)$; $g_eff(\tau)$; etc. To the end, it seems that the primary focus is the
intrinsic distribution $g(\tau)$; and relating this distribution with two other intrinsic
parameters: the reproduction number $\Ro$ and the Malthusian number $r$
}

The intrinsic distribution is not our primary focus. 
As we show in Figure 4, the intrinsic distribution overesitmates $\Rini$ as it does not account for spatial effects.
We tried to make clear that our goal is the initial forward distribution.

\revtext{
The authors have well demonstrated different sources of sampling bias
arising from contact tracing data, more noticeably, (i) length-bias due
to right-truncation because of retrospective measurement together with
different infection times during the course of the outbreak; (ii) due to the
network structure (e.g. the degree distribution and the degree excess (Line
408). Both sources, the right-truncation and the egocentric point of view
(Line 174), tend to give shorter on average observed intervals. These are
correct and important.
}

\revtext{
Any notion of bias is in the context against some "truth". The authors
proposed, and demonstrated by simulation, ways to adjust for the bias in
order to estimate this "truth". However, the authors equate the "true"
distribution to the intrinsic distribution $g(\tau)$ as if "obvious". It is not
obvious and this needs to be proven.
The intrinsic distribution $g(\tau)$ is an equilibrium distribution that
involves the equilibrium distribution of the infectious period (in SEIS,
SEIR, SEIRS settings). It is a renewal process concept and is only
valid when the system is under the equilibrium condition. It is not
obvious how it is related to measurements of "generation intervals"
when the system is not at equilibrium.
}

We now discuss this point in much more detail (and have discussed these points above). Briefly: we are interested in the initial forward distribution because this provides the \rR\ link. This approaches the intrinsic distribution in the case where mixing is homogeneous and the initial number infectious is small. We hope we have fixed all of the places where we failed to explain ourselves clearly.

@@

\revtext{
 Under suitable assumptions on homogeneous mixing, the intrinsic
distribution $g(\tau)$ measures the infectiousness within an infected individual without involving another individual. It is not obvious how it
is related to measures involving two individuals, no matter whether
they are prospectively or retrospectively measured.
}

\revtext{
Potentially observable events do not take place at the exact moment
when transmission occurs.
}

This is an important concern, and we have discussed this issue a bit more. We note that this is not always the case: for example, transmission is one of the most observable features in rabies dynamics. In Ebola, dates of transmission are often estimated. We are currently working to extend our framework to cases that explicitly treat other events as the observables.

### 2020-05-26 doc/response.tex

\revtext{
Lines 44-48: "... the observed generation-interval distribution can change...
...observed generation intervals refer to the time between actual infection
events." The actual infection events correspond to the exact moments of
passing the infection between a pair of infected-susceptible individuals.
Are they observable? In most cases, including contact tracing, observable
events are clinical (e.g. onset of symptoms). The relationship between
the interval defined by the time between actual infection events and the
interval defined by the time between clinical onsets is unclear. At least the
latter depends on the incubation period (of the infectee), in addition to the
latent period and the infectious period (of the infector). The incubation
period is defined as the time from infection to symptom onset, and is
different from the latent period such as in the SEIR model. This paper
does not make such distinction. It assumes that the time between actual
infection events "were observable" in a hypothetical way.
}

\revtext{
Suggestion: Re-phrase "observed generation intervals" or make some further
clarification on this. This is important because naive readers may tend
to take the face value and collect different kinds of observed data and
make inferences, such as that based on clinical events. This is potentially
misleading.
}

We appreciate this point. We now refer to them as \emph{realized} intervals.

\newtext{There are important distinctions to be made when defining generation-interval distributions at the population-level. The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.
On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.}

@@

We have also added the following sentence as per suggestions in previous comments:

\newtext{While an epidemic is ongoing, realized generation intervals can be measured by identifying who infected whom and can be aggregated to form a single distribution.
In practice, realized generation intervals is difficult to measure because it is difficult to observe the exact moment at which individuals become infected;
in most cases, observable events are clinical (e.g. onset of symptoms).
However, it is still possible to directly measure generation intervals through contact tracing in some cases (e.g., a rabid dog biting another dog \citep{hampson2009transmission}),.
Thus, the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic ---
we typically want to try to make inference based on all available data that have been gathered since the beginning of an epidemic.}

@@

\revtext{Lines 54-55: ``the backward generation interval...'' This is a key measurement in this paper along with the length-bias in relation to data observation scheme. Conceptually, there may also be a "forward" generation
interval. Both concepts were initially raised in Svensson (2007, ref.[21]
in the paper), who pointed out the difference in distribution between the
primary generation time as measured prospectively from the time of the
infection of the infector to the transmission to the infectee, and the secondary generation time as measured retrospectively from the time of the
transmission to the infectee to the infection of the infector.}

We now introduce the forward distribution as well.

\revtext{1. does that mean that your objective is to make unbiased inference on the
"forward", or the primary, generation time?}

Yes, we want to make inference on the \emph{initial} forward generation-interval distribution, rather than the intrinsic generation-interval distribution.
This is the distribution that correctly links the initial values of $r$ and \Rini.
We have tried to make this clear and no longer use the term ``effective'' generation-interval distribution.

@@

\newtext{Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.}

\revtext{
The "the backward generation interval (or the secondary generation time)" is statistically sound. An infectee, before being infected, has been exposed to many infectors in a prevalent cohort and
could be independently infected by any of them. Each infectee is only associated with a single backward generation time whose distribution arises
from an independent competing risk framework. At the population level,
all backward measured intervals are independent random variables (but
not necessarily identically distributed). On the other hand, the "forward
generation interval" is problematic. A typical infected individual may
infect several contacts, but not independently if the infectious period is
random. Therefore, it is not appropriate to consider that the times to
transmission of the pathogen to its infectees as an iid. sample. This can
be shown in Exercise 3.5 (f) of Yan and Chowell (2019). At the population level, should such data can be observed, there might exist correlated
intervals arising from the same infector, something to ponder about.
}

We appreciate this point and have tried not to imply strong assumptions, and also taken care to empirically validate our statistical approaches.

\swp{??? What do you mean by ``also taken care to empirically validate our statistical approaches''}

@@

\revtext{
Lines 59, 114-115, 129, etc.: There is a distinction between "truncation"
and "censoring" in survival analysis. In the current context (e.g. contact tracing), when the inclusion criteria is conditioning on the infection
event of the infectee and looking backward to ascertain the infection time
of its infector, it is right-truncation. It leads to over-sampling short durations (as stated in Line 129). I suggest to use "right-truncation" instead
of "censoring" throughout the paper, wherever it appears.
}

Done.

\revtext{
Lines 121 and 123: The backward generation-interval distribution $b_t(\tau)$ and
the "censored" (right-truncated?) generation-interval distribution $c_t(\tau)$
should be clearly defined and expressed (e.g. in terms of conditional distribution) before expressing the relationship (6).
}

We now define both backward and forward generation-interval distributions more appropriately in the text.
We now use $a_t$ to refer to the ``aggregated'' generation-interval distribution (all realized generation intervals until certain time point) and do not use $c_t$.

@@

\revtext{
When stating that ``$\RR$'' transcends to the final size of the epidemic (Line
37), do the authors actually mean $\Ro$; which does not depend on time?
}

Yes. We now make a clear distinction between $\RR$ and $\Ro$.

\revtext{
When talking about the final size, one considers that the system is at a
disease-free equilibrium. When using the final size equation to calculate a
value for the reproduction number, the calculated value does not depend
on time. If there are behavior changes and public health interventions
throughout the outbreak, the calculated value $\RR_{final}$ is usually smaller
than $\Ro$ (defined at another disease-free equilibrium at the very moment
when an infector was seeded). So now we have three reproduction numbers, $\Ro$, a time-dependent $\RR = \Ro S(t)$ and $\RR_{final}$: Need to clarify the differences and reveal the hidden assumptions when discussing (and using)
each of them.
}

We are interested in the initial value of $\RR$, which we now refer to as $\Rini = \Ro S(0)$. We have tried to make this clear throughout the manuscript.

@@

While we agree with the reviewer that the intrinsic generation-interval can be interpreted an equilibrium distribution, this only holds under homogeneous mixing assumption. 
As we show in this paper, when there is spatial structure, the realized generation-interval distributions differ from the intrinsic one. 
Even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egocentric effect.
Therefore, the realized generation intervals of a single, primary infector in an otherwise susceptible population will not follow the intrinsic generation-interval distribution. We do mention in the manuscript that the intrinsic distribution can be interpreted as an equilibrium distribution but do not want to over-emphasize this perspective, since we're interested in realized distributions on a spatial network.
The reviewer also presented a series of formulas on the mean generation interval for special cases of the SEIR model; these are interesting, but not a direction that we want to go with this version of the paper.

\revtext{
Expression (13) is the convolution (the sum of) the exponentially
distributed latent period with mean $1/\sigma$ and another exponential distribution, with mean $1/(\gamma+\lambda)$; as written in Lines 188-189. It is also equivalent
to the homogeneous mixing situation, in which, individuals during their
infectious periods are isolated (with rate $\lambda$) so that the "effective" duration
of the infectious period is exponentially distributed with mean $1/(\gamma + \lambda)$:
Any comments from the authors on this coincidence?
}

We explain this in the main text:

\newtext{In this case, with constant transmission rate during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$). 
Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.}

@@

\revtext{
It is more convenient to use the Laplace transforms as a tool.
}

The reviewers discussion of and suggestions about Laplace transformations are interesting, but we think beyond the scope of this paper. Our experience has been that Laplace transformations and moment generating functions are as likely to obscure the key concepts underlying disease dynamics as to clarify them.

\revtext{How to assess if the gamma distribution is a good choice as a
model? Run some kind of goodness-of-fit tests?}

These are hard issues, but beyond the scope of this paper. 
Using exponentials is routine in many applications. 
Using gammas (or Erlangs) generalizes this, and is also commonly done. 
The gamma distribution has been previously tested for many diseases. 
In our case, Figure 4 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as potential caveats in assuming a shape of the generation-interval distribution in Discussion.

\newtext{We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity;
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estiamte of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of \Rini (Appendix A.4).}

@@

\revtext{How does the likelihood
function look like? Since the "observed" generation-intervals (assuming
observable) do not arise from a designed random experiment, but observational data, I expect the likelihood function should reflect the observation
scheme (e.g. retrospective)? Please be explicit.
}

The population-level method does not take into account the observation scheme or the infection process. 
We simply fit the gamma distribution to all realized generation intervals (i.e., the aggregated distribution).
We have clarified the text:

\newtext{The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom) \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau).
\end{equation}
In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).}

@@

\revtext{
How exactly is $\RR$ (and, is it \Ro or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref. [54] in the paper)?
}

We estimate $\Rini$ via Euler-Lotka. We have tried to make this clear throughout the main text.

\revtext{
How close is the "related approach" in White and Pagano
(2007, ref. [54] in the paper) related to the current paper? Did they use the
Euler-Lotka equation at all? Did they define and measure the generation intervals that were consistent with the generation-intervals discussed in
the current paper?
}

We had initially thought their approach is similar due to their use of Poisson likelihood.
We realized that their approach is quite different from the goal of our paper and have removed the sentence.
They do not use Euler-Lotka or measure generation intervals.

@@

\revtext{
The authors use a logistic function with a scale parameter $r$,
the carrier capacity parameter $K$ and the initial condition $c_0$ to model the
cumulative incidence $c(t)$: Assuming, (theoretically), the infection events
are observable. It is inaccurate to claim that r is the exponential growth
rate
}

We have added the following justification to the text:

\newtext{
We estimate the initial exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}:
\begin{equation}
c(t) = \frac{K}{1 + \left[(K/c_0)-1\right]e^{-rt}}.
\end{equation}
While the exponential growth rate (i.e., the rate of change in log incidence) of the logistic function changes throughout an epidemic,
we focus strictly on estimating the \emph{initial} exponential growth rate (when $t \to -\infty$).
The method of estimating the initial exponential growth rate by fitting a logistic curve has been previously validated against simulations of stochastic comparmental models \citep{ma2014estimating}.
}

@@

\revtext{Minor points:}

\revtext{Figure 1: the average latent period is 11.4 days and the average infectious period is 5 days. However, these seem to be switched in Figures 2, 3 \& 4. Why?}

Thank you for pointing this out. We made typographical errors in figure captions. We have corrected this.

\revtext{Line 145: "acquantanceships" $\rightarrow$ "acquaintanceships"}
